Figures and results for main manuscript

### 2020-12-17 analysis/reports/appendix1_filtering.Rmd
appendices

We filter datasets in two stages. First, **prior to trying to sample from the feasible set**, we remove communities that have a combined S and N too high for us to sample the feasible set. We also create a sub-sample of the FIA dataset, because it has so many communities (~100,000) and so many of them are extremely small (~90,000 with fewer than 10 species). Sampling all 100,000 communities overwhelms our computational pipeline, so we create a subsample of 20,000 communities comprising all communities with more than 10 species (approx. 10,000) and 10,000 randomly selected communities with 3-10 species. Second, **after sampling but before we aggregate results across communities**, we remove communities that are problematic for a number of more nuanced reasons. This includes having only 1 possible SAD (S = 1, N = S, or N = S + 1); mixed relative abundance and count data (for the Misc. Abundance Database). We also filter to communities whose sampled feasible sets yield more than 20 unique values for skewness/evenness. We do this because, if there are fewer than 20 unique values in the comparison vector, it's impossible to be in the 5th or 95th percentile. Finally, we remove communities with only 2 species from analyses for skewness, because  `e1071::skewness()` always = 0 if S = 2.

@@

- Misc. Abund includes datasets reported as relative abundance in addition to count data. We don't want any communities without counts, so we filter out records where abund = 0. *While preparing this report, RMD noticed that some communities have a __mixture__ of relative abundance and count data. These communities were sampled, but RMD does not believe it is appropriate to compare them to count-data-only communities. We therefore remove these 6 in the second, post-sampling, phase of filtering.*

@@

We remove communities with only one possible SAD (N = S, N = S + 1, or S = 1) and five communities in the Misc. Abund Database that have mixed count and relative abundance data (there are more than 5 with a mixture, but only 5 that have both the mixture AND make it through the previous filters).

@@

```{r}
miscabund_problems <- as.character(c(144, 145, 145, 146, 147, 178, 446))

all_di_filtered <- all_di_filtered %>%
  mutate(is.miscabund.problem = ifelse(dat == "misc_abund",
                                       ifelse(site %in% miscabund_problems, T, F), F))

filter(all_di_filtered, is.miscabund.problem)

all_di_filtered <- all_di_filtered %>%
  filter(!is.miscabund.problem) %>%
  ungroup() %>%
  select(-is.miscabund.problem)

@@

all_di_filtered %>%
  filter(simpson_unique > 20) %>%
  group_by(dat) %>%
  summarize(sites_for_evenness = dplyr::n())

all_di_filtered %>%
  filter(skew_unique > 20, s0 > 2) %>%
  group_by(dat) %>%
  summarize(sites_for_skewness = dplyr::n())

>>>>

We filter datasets in two stages. First, **prior to trying to sample from the feasible set**, we remove communities that have a combined S and N too high for us to sample the feasible set. We also create a sub-sample of the FIA dataset, because it has so many communities (~100,000) and so many of them are extremely small (~90,000 with fewer than 10 species). Sampling all 100,000 communities overwhelms our computational pipeline, so we create a subsample of 20,000 communities comprising all communities with more than 10 species (approx. 10,000) and 10,000 randomly selected communities with 3-10 species. Second, **after sampling but before we aggregate results across communities**, we remove communities that are problematic for a number of more nuanced reasons. This includes having only 1 possible SAD (S = 1, N = S, or N = S + 1). We also filter to communities whose sampled feasible sets yield more than 20 unique values for skewness/evenness. We do this because, if there are fewer than 20 unique values in the comparison vector, it's impossible to be in the 5th or 95th percentile. Finally, we remove communities with only 2 species from analyses for skewness, because  `e1071::skewness()` always = 0 if S = 2.

@@

- Misc. Abund includes datasets reported as relative abundance in addition to count data. We don't want any communities without counts, so we filter out records where abund = 0.

@@

We remove communities with only one possible SAD (N = S, N = S + 1, or S = 1).

@@

@@

all_di_filtered %>%
  filter(simpson_unique > 20) %>%
  group_by(dat) %>%
  summarize(sites_for_evenness = dplyr::n()) %>%
  ungroup() %>%
  mutate(total_sites_for_evenness = sum(sites_for_evenness))

all_di_filtered %>%
  filter(skew_unique > 20, s0 > 2) %>%
  group_by(dat) %>%
  summarize(sites_for_skewness = dplyr::n()) %>%
  mutate(total_sites_for_skewness = sum(sites_for_skewness))

### 2020-12-17 analysis/reports/appendix2_selfsimilarity.Rmd
appendices

---
title: "Self-similarity of the elements of the feasible set"
author: Renata Diaz
date: "`r Sys.Date()`"
output: 
    word_document:
       df_print: kable
---

>>>>

---
title: "Self-similarity of the elements of the feasible set"
author: Renata Diaz
date: "`r Sys.Date()`"
output: 
    github_document:
       df_print: kable
---

### 2020-12-17 analysis/reports/manuscript_main.Rmd
misc abund rabbit hole
appendices

all_di <- all_di %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)),
         log_nsamples = log(nsamples),
         log_s0 = log(s0),
         log_n0 = log(n0)) %>%
  filter(n0 > s0,
         s0 > 1,
         !singletons,
         dat %in% c("bbs", "fia_short", "fia_small", "gentry", "mcdb", "misc_abund_short")) %>%
  mutate(dat = ifelse(grepl(dat, pattern = "fia"), "fia", dat),
         dat = ifelse(dat == "misc_abund_short", "misc_abund", dat)) %>%
  mutate(Dataset = dat,
    Dataset = ifelse(Dataset == "fia", "Forest Inventory and Analysis", Dataset),
        Dataset = ifelse(Dataset == "bbs", "Breeding Bird Survey", Dataset),
        Dataset = ifelse(Dataset == "mcdb", "Mammal Community DB", Dataset),
        Dataset = ifelse(Dataset == "gentry", "Gentry", Dataset),
        Dataset = ifelse(Dataset == "misc_abund", "Miscellaneous Abundance DB", Dataset))
```

# Final dataset in S and N space

```{r final dataset s and n space}

ggplot(all_di, aes(x = log_s0, y = log_n0, color = Dataset)) +
  geom_point(alpha = .1) +
  geom_point(data = filter(all_di, log_nparts > log(50)), alpha = .4) +
  theme_bw() +
  scale_color_viridis_d() +
  ggtitle("Communities by S and N") +
  xlab("Log species richness (S)") +
  ylab("Log total abundance (N)")

ggplot(all_di, aes(x = log_s0, y = log_n0, color = Dataset)) +
  geom_point(alpha = .05) +
  geom_point(data = filter(all_di, log_nparts > log(50)), alpha = .4) +
  theme_bw() +
  scale_color_viridis_d() +
  ggtitle("Communities by S and N") +
   facet_wrap(vars(Dataset), scales = "free") +
  theme(legend.position = "bottom")
  # geom_vline(xintercept = log(40)) +
  # geom_hline(yintercept = log(1000))

@@

# Illustrations of 95% interval

To show the 95% interval, we need to load the distribution of shape metric values from the samples from the feasible set for a few communities.

# Skewness and evenness histograms by dataset

```{r first hists}

ggplot(filter(all_di, s0 > 2, skew_unique > 20), aes(skew_percentile_excl)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = 95, color = "red") +
  theme_bw() +
  facet_wrap(vars(Dataset), scales = "free_y") + ggtitle("Skewness, all sites") +
  xlab("Percentile rank of observed skewness value relative to feasible set") +
  ylab("Number of communities")

ggplot(filter(all_di, simpson_unique > 20), aes(simpson_percentile)) + 
 geom_histogram(bins = 100) +
  geom_vline(xintercept = 5, color = "red") +
  theme_bw() +
  facet_wrap(vars(Dataset), scales = "free_y") + ggtitle("Evenness, all sites") +
  xlab("Percentile rank of observed evenness value relative to feasible set") +
  ylab("Number of communities")

@@

# Proportion of sites with highly skewed or uneven SADs

```{r results table}

overall_skew_results <- all_di %>%
  filter(s0 > 2, skew_unique > 20) %>%
  mutate(high_skew = skew_percentile_excl > 95) %>%
  group_by(Dataset) %>%
  summarize(proportion_skew_high = mean(high_skew),
            nsites_skew = dplyr::n())

overall_simpson_results <- all_di %>%
  filter(simpson_unique > 20) %>%
  mutate(low_even = simpson_percentile < 5) %>%
  group_by(Dataset) %>%
  summarize(proportion_even_low = mean(low_even),
            nsites_even = dplyr::n())

overall_results <- left_join(overall_skew_results, overall_simpson_results) %>%
  rename("Proportion of communities with skewness above 95th percentile" = proportion_skew_high,
         "Number of communities analyzed for skewness" = nsites_skew,
         "Proportion of communities with evenness below 5th percentile" = proportion_even_low,
         "Number of communities analyzed for evenness" = nsites_even)

overall_results

overall_skew_results2 <- all_di %>%
   mutate(Dataset = ifelse(dat == "fia", "Forest Inventory and Analysis", "Other datasets")) %>%
  filter(s0 > 2, skew_unique > 20) %>%
  mutate(high_skew = skew_percentile_excl > 95) %>%
  group_by(Dataset) %>%
  summarize(proportion_skew_high = mean(high_skew),
            nsites_skew = dplyr::n())

overall_simpson_results2 <- all_di %>%
  filter(simpson_unique > 20) %>%
   mutate(Dataset = ifelse(dat == "fia", "Forest Inventory and Analysis", "Other datasets")) %>%
  mutate(low_even = simpson_percentile < 5) %>%
  group_by(Dataset) %>%
  summarize(proportion_even_low = mean(low_even),
            nsites_even = dplyr::n())

overall_results2 <- left_join(overall_skew_results2, overall_simpson_results2) %>%
   rename("Proportion of communities with skewness above 95th percentile" = proportion_skew_high,
         "Number of communities analyzed for skewness" = nsites_skew,
         "Proportion of communities with evenness below 5th percentile" = proportion_even_low,
         "Number of communities analyzed for evenness" = nsites_even)

overall_results2

@@

# 95 intervals by size of FS

```{r 95 intervals}

ggplot(filter(all_di, s0 > 2, skew_unique > 20), aes(x = log_nparts, y= skew_95_ratio_1t)) +
  geom_point(alpha = .1) +
  theme_bw() +
  xlim(0, 100) +
  ylim(0,1)+
  xlab("Log number of elements in the feasible set") +
  ylab("Narrowness of the distribution of sampled skewness values")

ggplot(filter(all_di, simpson_unique > 20), aes(x = log_nparts, y= simpson_95_ratio_1t)) +
  geom_point(alpha = .1) +
  theme_bw() +
  xlim(0, 100) +
  ylim(0, 1)+
  xlab("Log number of elements in the feasible set") +
  ylab("Narrowness of the distribution of sampled evenness values")
```

# 95 intervals by dataset

```{r}

ggplot(filter(all_di, s0 > 2, skew_unique > 20), aes(x= skew_95_ratio_1t)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(vars(Dataset), scales = "free_y") +
    xlab("Narrowness of the distribution of sampled skewness values") +
  ylab("Number of communities")

ggplot(filter(all_di, simpson_unique > 20), aes(x = simpson_95_ratio_1t)) +
geom_histogram() +
  theme_bw() +
facet_wrap(vars(Dataset), scales = "free_y") +
    xlab("Narrowness of the distribution of sampled evenness values") +
  ylab("Number of communities")

@@

>>>>

all_di <- all_di %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)),
         log_nsamples = log(nsamples),
         log_s0 = log(s0),
         log_n0 = log(n0)) %>%
  filter(n0 != s0,
         s0 != 1,
         !singletons,
         n0 != (s0 + 1)) %>%
  mutate(dat = ifelse(grepl(dat, pattern = "fia"), "fia", dat),
         dat = ifelse(dat == "misc_abund_short", "misc_abund", dat)) %>%
  mutate(Dataset = dat,
         Dataset = ifelse(Dataset == "fia", "FIA", Dataset),
         Dataset = ifelse(Dataset == "bbs", "Breeding Bird Survey", Dataset),
         Dataset = ifelse(Dataset == "mcdb", "Mammal Communities", Dataset),
         Dataset = ifelse(Dataset == "gentry", "Gentry", Dataset),
         Dataset = ifelse(Dataset == "misc_abund", "Misc. Abundance", Dataset))

@@

# Illustrations of 95% interval (Figure 1)

To show the 95% interval, we need to load the distribution of shape metric values from the samples from the feasible set for a few communities. See rov_metric.md.

```{r, fig.dim = c(7,7)}
library(drake)

db <- DBI::dbConnect(RSQLite::SQLite(), here::here("analysis", "drake", "drake-cache-net.sqlite"))
cache <- storr::storr_dbi("datatable", "keystable", db)
cache$del(key = "lock", namespace = "session")

net_summary <- readd(all_di_summary, cache = cache)

net_summary <- net_summary %>%
  mutate(log_nparts = log(gmp:::as.double.bigz(nparts)))

example_fs <- readd(fs_s_44_n_13360, cache = cache)

example_di <- readd(di_fs_s_44_n_13360, cache =cache)

example_fs <- example_fs %>%
  left_join(example_di) %>%
  left_join(net_summary)

example_fs2 <- readd(fs_s_7_n_71, cache = cache)

example_di2 <- readd(di_fs_s_7_n_71, cache =cache)

example_fs2 <- example_fs2 %>%
  left_join(example_di2) %>%
  left_join(net_summary)

example_fs3 <- readd(fs_s_4_n_34, cache = cache)

example_di3 <- readd(di_fs_s_4_n_34, cache =cache)

example_fs3 <- example_fs3 %>%
  left_join(example_di3) %>%
  left_join(net_summary)

breadth_plots <- list(

  ggplot(example_fs3, aes(rank, abund, group = sim, color = skew)) +
    geom_line(alpha = .25) +
    theme_bw() +
    scale_color_viridis_c(option = "plasma", end = .8) +
    ggtitle("Small community", subtitle = paste0("S = ", (example_fs3$s0), "; N = ", (example_fs3$n0[1]))) +
    theme(legend.position = "right"),

  ggplot(example_fs3, aes(skew, ..ndensity..)) +
    geom_density() +
    geom_histogram(bins = 50) +
    theme_bw() +
    geom_vline(xintercept = c(example_fs3$skew_95[1], example_fs3$skew_min[1]), color = "red") +
    ggtitle("", subtitle = paste0("Breadth index (skewness): ", round((example_fs3$skew_95_ratio_1t[1]), 2))),

  ggplot(example_fs2, aes(rank, abund, group = sim, color = skew)) +
    geom_line(alpha = .1) +
    theme_bw() +
    scale_color_viridis_c(option = "plasma", end = .8) +
    ggtitle("Medium community", subtitle = paste0("S = ", (example_fs2$s0), "; N = ", (example_fs2$n0[1]))) +
    theme(legend.position = "right"),

  ggplot(example_fs2, aes(skew, ..ndensity..)) +
    geom_density() +
    geom_histogram(bins = 50) +
    theme_bw() +
    geom_vline(xintercept = c(example_fs2$skew_95[1], example_fs2$skew_min[1]), color = "red") +
    ggtitle("", subtitle =  paste0("Breadth index (skewness): ", round((example_fs2$skew_95_ratio_1t[1]), 2))),

  ggplot(example_fs, aes(rank, abund, group = sim, color = skew)) +
    geom_line(alpha = .1) +
    theme_bw() +
    scale_color_viridis_c(option = "plasma", end = .8) +
    ggtitle("Large community", subtitle = paste0("S = ", (example_fs$s0), "; N = ", (example_fs$n0[1]))) +
    theme(legend.position = "right"),

  ggplot(example_fs, aes(skew, ..ndensity..)) +
    geom_density() +
    geom_histogram(bins = 50) +
    theme_bw() +
    geom_vline(xintercept = c(example_fs$skew_95[1], example_fs$skew_min[1]), color = "red") +
    ggtitle("", subtitle =  paste0("Breadth index (skewness): ", round((example_fs$skew_95_ratio_1t[1]), 2)))

)

fig_1 <- gridExtra::grid.arrange(grobs = breadth_plots, ncol = 2)

plot(fig_1)

DBI::dbDisconnect(db)
rm(cache)

@@

# Skewness and evenness histograms by dataset (Figure 2)

```{r first hists, fig.dim = c(7,7)}
fig_2 <- gridExtra::grid.arrange(grobs = list(

  ggplot(filter(all_di, s0 > 2, skew_unique > 20), aes(skew_percentile_excl)) +
    geom_histogram(bins = 100) +
    geom_vline(xintercept = 95, color = "red") +
    theme_bw() +
    facet_wrap(vars(Dataset), scales = "free_y") + ggtitle("Skewness") +
    xlab("Percentile rank of observed value relative to feasible set") +
    ylab("Number of communities"),

  ggplot(filter(all_di, simpson_unique > 20), aes(simpson_percentile)) + 
    geom_histogram(bins = 100) +
    geom_vline(xintercept = 5, color = "red") +
    theme_bw() +
    facet_wrap(vars(Dataset), scales = "free_y") + ggtitle("Evenness") +
    xlab("Percentile rank of observed value relative to feasible set") +
    ylab("Number of communities")
))

plot(fig_2)

@@

# Breadth index by dataset (Figure 3)

```{r, fig.dim = c(7,7)}

fig_3 <- gridExtra::grid.arrange(grobs = list(
  ggplot(filter(all_di, s0 > 2, skew_unique > 20), aes(x= skew_95_ratio_1t)) +
    geom_histogram() +
    theme_bw() +
    facet_wrap(vars(Dataset), scales = "free_y") +
    xlab("Breadth index") +
    ylab("Number of communities") +
    ggtitle("Skewness"),

  ggplot(filter(all_di, simpson_unique > 20), aes(x = simpson_95_ratio_1t)) +
    geom_histogram() +
    theme_bw() +
    facet_wrap(vars(Dataset), scales = "free_y") +
    xlab("Breadth index") +
    ylab("Number of communities") +
    ggtitle("Evenness")
))

plot(fig_3)

@@

# Comparison of FIA and comparably sized communities (Figure 4)

```{r, fig.dim = c(7,7)}

fia_max_s = max(filter(all_di, dat == "fia")$s0)
fia_max_n = max(filter(all_di, dat == "fia")$n0)

fia = filter(all_di, dat == "fia")

small_di <- all_di %>%
  filter(s0 <= fia_max_s,
         n0 <= fia_max_n) %>%
  mutate(fia_yn = ifelse(dat == "fia", "fia", "other datasets"),
         right_corner = FALSE)

for(i in 1:nrow(small_di)) {
  if(small_di$dat[i] != "fia") {
    fia_match_s0 = filter(fia, s0 >= small_di$s0[i])

    max_n0_for_this_s0 = max(fia_match_s0$n0)

    small_di$right_corner[i] = small_di$n0[i] > max_n0_for_this_s0

  }
}

small_di <- small_di %>%
  filter(!right_corner)
not_fia <- filter(small_di, fia_yn != "fia")
fia = filter(small_di, dat == "fia")

small_di_s_n <- small_di %>%
  select(s0, n0) %>%
  distinct() %>%
  mutate(in_fia = FALSE,
         in_not_fia = FALSE)

for(i in 1:nrow(small_di_s_n)) {

  this_s0 = small_di_s_n$s0[i]
  this_n0 = small_di_s_n$n0[i]

  fia_matches = filter(fia, s0 == this_s0, n0 == this_n0)
  not_fia_matches = filter(not_fia, s0 == this_s0, n0 == this_n0)

  small_di_s_n$in_fia[i] = nrow(fia_matches) > 0 
  small_di_s_n$in_not_fia[i] = nrow(not_fia_matches) > 0

}

small_di_s_n <- filter(small_di_s_n, in_fia, in_not_fia)

set.seed(1977)

subsamples <- list()

for(i in 1:nrow(small_di_s_n)) {

  this_s0 = small_di_s_n$s0[i]
  this_n0 = small_di_s_n$n0[i]

  fia_matches = filter(fia, s0 == this_s0, n0 == this_n0)

  not_fia_matches = filter(not_fia, s0 == this_s0, n0 == this_n0)

  n_to_draw = min(nrow(fia_matches), nrow(not_fia_matches))

  fia_draw = sample.int(n = nrow(fia_matches), size = n_to_draw, replace = F)
  not_fia_draw = sample.int(n = nrow(not_fia_matches), size = n_to_draw, replace = F)

  subsamples[[i]] <- (bind_rows(fia_matches[fia_draw,], not_fia_matches[not_fia_draw, ]))

}

sub_di <- bind_rows(subsamples) %>%
  mutate(Dataset = ifelse(fia_yn == "fia", "FIA", "Other datasets"))

fig_4 <- gridExtra::grid.arrange(grobs = list(
  ggplot(filter(sub_di, skew_unique > 20), aes(skew_95_ratio_1t)) +
    geom_histogram()+
    theme_bw() +
    facet_wrap(vars(Dataset), scales = "free_y")  +
    xlab("Breadth index") +
    ylab("Number of communities") +
    ggtitle("Breadth index", subtitle = "Skewness"),

  ggplot(filter(sub_di, simpson_unique > 20), aes(simpson_95_ratio_1t)) +
    geom_histogram()+
    theme_bw() +
    facet_wrap(vars(Dataset), scales = "free_y")  +
    xlab("Breadth index") +
    ylab("") +
    ggtitle("", subtitle = "Evenness"),

  ggplot(filter(sub_di, skew_unique > 20), aes(skew_percentile_excl)) +
    geom_histogram()+
    theme_bw() +
    facet_wrap(vars(Dataset), scales = "free_y")  +
    xlab("Percentile rank") +
    ylab("Number of communities") +
    ggtitle("Percentile rank", subtitle = "Skewness"),

  ggplot(filter(sub_di, simpson_unique > 20), aes(simpson_percentile)) +
    geom_histogram()+
    theme_bw() +
    facet_wrap(vars(Dataset), scales = "free_y")  +
    xlab("Percentile rank") +
    ylab("")+
    ggtitle("", subtitle = "Evenness")), ncol = 2)

plot(fig_4)

```{r}
pdf("figure_1.pdf")
plot(fig_1)
dev.off()

pdf("figure_2.pdf")
 plot(fig_2)
 dev.off()

pdf("figure_3.pdf")
plot(fig_3)
dev.off()

pdf("figure_4.pdf")
 plot(fig_4)
dev.off()
```
